---
title: "Course Project 1"
author: "John Magcamit"
date: '2022-07-26'
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Loading and preprocessing the data
```{r, message=FALSE, warning=FALSE}
library(dplyr)
df <- read.csv("activity.csv", sep = ",", colClasses = c("numeric", "Date", "numeric"))
df$date <- as.POSIXct(df$date, "%Y%m%d")
```
```{r}
head(df, 10)
```

## Total number of steps taken each day
Total number of steps each day
```{r}
df1 <- df %>% group_by(date) %>% summarize(TotalSteps = sum(steps))
head(df1, 10)
```
Histogram of Total Steps per Day
```{r}
hist(df1$TotalSteps, xlab="Number of steps taken each day", col="green", breaks = length(df1$date)/4, main = "Histogram of Total Steps per Day")
```

Mean and median number of steps taken each day (excluding NA)

```{r}
averageSteps <- mean(df1$TotalSteps, na.rm = TRUE)
medianSteps <- median(df1$TotalSteps, na.rm = TRUE)
```

```{r, echo=FALSE}
return(paste("Average steps: ",averageSteps))
return(paste("Median steps: ",medianSteps))
```
## Average daily activity pattern

```{r}
df2 <- df %>% group_by(interval) %>% summarize(aveSteps=mean(steps, na.rm=TRUE))
head(df2, 10)
```
Time series plot
```{r}
plot(df2$interval, df2$aveSteps, type="l", xlab="5-Minute Interval", ylab="Average Number of Steps", main = "Time series plot of the 5-Minute Interval and average number of steps")  
```

5-minute interval, on average across all the days in the dataset, contains the maximum number of steps

```{r}
maxSteps <- which.max(df2$aveSteps)
df2[maxSteps,2]
```
Maximum steps for 5 minute interval is 206

## Imputing missing values
Total number of missing values in the dataset
```{r}
sum(is.na(df$steps))
```

```{r}
newdf <- df
newdf$steps[which(is.na(newdf$steps) & newdf$interval==df2$interval)] <- df2$aveSteps
head(newdf, 10)
```

Histogram of the total number of steps taken each day after missing values are imputed
```{r}
df3 <- newdf %>% group_by(date) %>% summarize(TotalSteps = sum(steps))
hist(df3$TotalSteps, xlab="Number of steps taken each day", col="green", breaks = length(df1$date)/4, main = "Histogram of Total Steps per Day")
```

Mean and median of total number of steps per day
```{r}
mean(df3$TotalSteps)
median(df3$TotalSteps)
```

##  Differences in activity patterns between weekdays and weekends

Using newdf as dataset with the filled-in missing values
```{r}
df4 <- newdf
weekend <- c("Saturday", "Sunday")
df4$day <- weekdays(df4$date)
df4$day[df4$day=="Saturday" | df4$day=="Sunday"] <- "weekend"
weekendDF <- df4 %>% filter(df4$day=="weekend") %>% group_by(interval) %>% summarize(aveSteps=mean(steps))
weekdayDF <- df4 %>% filter(df4$day!="weekend") %>% group_by(interval) %>% summarize(aveSteps=mean(steps))
```

Plotting the weekend vs weekday average steps/day
```{r}
par(mfrow=c(2,1), mar=c(4,4,2,2))
plot(weekendDF$interval, weekendDF$aveSteps, type="l", main="Weekend", col="blue", xlab="5-Minute Intervals", ylab="Average Steps per Day")
plot(weekdayDF$interval, weekdayDF$aveSteps, type="l", main="Weekday", col="red", xlab="5-Minute Intervals", ylab="Average Steps per Day")

```